Parkinson's Disease cluster analysis

### 2016-05-23 writeup/manuscript/manuscript.tex
More minor tweaks to corrected manuscript

% Use times font
\usepackage{mathptmx}
\usepackage{lmodern}
\usepackage{times}

@@

$k$-means was used for cluster analysis. We performed two analyses on the
patients in the dataset: the first ``domains clustering'' on the nine aggregate nonmotor symptom
domains and the four motor symptoms, and the second ``symptoms clustering'' on the 30 individual
nonmotor symptoms collected with the four motor symptoms. Complete-linkage agglomerative
hierarchical clustering on the 30 nonmotor symptoms was also performed to how the symptoms group
together.

@@

Various formal measures were used to determine the optimal number of clusters for the dataset. The
optimal $k$ according to the Gap Statistic and the 1-standard-error method suggested by
Tibshirani\cite{tibshirani01gap} was $k = 4$ (Supplementary Material Figure~\ref{fig:gap}). Other cluster validation
methods (within sum of squares (WSS) scree plot, minimum average silhouette width) suggested $k =
2, 3, 4$, where $k = 2, 3$ simply divided the data uninformatively into groups with varying levels
of overall PD severity. Thus $k = 4$ was selected to offer a good blend of model fit,
informativeness, and parsimony.

@@

To compare the clusterings, we visualized cluster alignment with stacked barplots, and computed the adjusted rand index\cite{hubert1985} (ARI) and
Rosenberg's $v$-measure.\cite{rosenberg07vmeasure} to evaluate similarity between the two
clusterings. Both measures range from 0 (no similarity) to 1 (identical), and respectively take a
contingency table and information-theoretic approach to measuring clustering similarity.

@@

$k$-means clustering performed on the 30 individual nonmotor symptoms found similar patterns to the
first clustering (Table~\ref{tab:nms30}, heatmap in Figure~\ref{fig:nms30-hm}). Means of all
symptoms were found to differ across clusters except for PD onset and tremor.

@@

 The two clusterings were somewhat similar ($\text{ARI} = 0.46$; $\text{$v$-measure} = 0.43$).  To
 help compare the alignment of the two clusterings, we denoted the four clusters in the domains
 clustering as $D_1, D_2, D_3, D_4$, and the clusters in the symptoms clustering as $S_1, S_2, S_3,
 S_4$.
Alignment of the clusters is visualized in Figure~\ref{fig:align}. While clusters $D_1/S_1$ and
$D_4/S_4$
aligned well with each other, main differences in the clusterings occurred in the mixing of
clusters 2 and 3; for example, $D_3$ (motor-dominant) was somewhat
evenly split among $S_1$ (mild) and $S_3$ (average).

@@

Most variables had little to no correlation with PD duration (Supplementary Material Figure~\ref{fig:corr}). Variables
with the highest correlation include symptoms urinary frequency, swallowing, and drooling; domains
miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots for CISI Total,
Tremor, Anxiety, and Depression appear in Supplementary Material Figure~\ref{fig:long4}.

@@

Our longitudinal analysis gives more insight into the clusters found in the previous analyses.
According to Figure~\ref{fig:corr}, most symptoms are uncorrelated with PD duration, especially
Mood/apathy symptoms (anxiety, depression) and tremor.
The differences in disease progression for each cluster can be seen by the corresponding graphs in
Figure~\ref{fig:long4}. Cluster 2 (Nonmotor-Dominant) starts at higher
scores for anxiety and depression, and actually decreases with increasing PD duration, thus indicating that these patients' subtype can be determined
early after PD onset from the depressive symptom score. Similarly, when examining cluster 3
(Motor-dominant), the mean tremor score is substantially higher from PD onset.
Interestingly, cluster 4 (Severe) generally starts at lower tremor and motor scores during disease,
but then rises sharply, exceeding other clusters. More evidence that tremor is a unique motor
symptom is located in Figure~\ref{fig:hc}, where it is the most distant symptom from all other
symptoms.

@@

\begin{suppfigure*}[p]
  \centering
  \includegraphics[width=0.8\linewidth]{long4-d.pdf}
  \caption{Selected symptoms for 901 patients plotted against PD duration. Smoothed loess curves for each cluster
  are drawn in their respective colors. The black curve is the curve for the entire population, and
the global mean score is marked with a dotted line.}
  \label{fig:long4}
\end{suppfigure*}

>>>>

% Use times font
\usepackage{mathptmx}
\usepackage{lmodern}
\usepackage{times}
\usepackage{soul} % For highlighting

@@

$k$-means was used for cluster analysis. We performed two analyses on the
patients in the dataset: the first ``domains clustering'' on the nine aggregate nonmotor symptom
domains and the four motor symptoms, and the second ``symptoms clustering'' on the 30 individual
nonmotor symptoms collected with the four motor symptoms. Complete-linkage
hierarchical agglomerative clustering on the 30 nonmotor symptoms was also performed to how the symptoms group
together.

@@

Various formal measures were used to determine the optimal number of clusters for the dataset. The
optimal $k$ according to the Gap Statistic and the 1-standard-error method suggested by
Tibshirani\cite{tibshirani01gap} was $k = 4$ (Supplementary Figure~\ref{fig:gap}). Other cluster validation
methods (within sum of squares scree plot, minimum average silhouette width) suggested $k =
2, 3, 4$, where $k = 2, 3$ simply divided the data uninformatively into groups with varying levels
of overall PD severity. Thus $k = 4$ was selected to offer a good blend of model fit,
informativeness, and parsimony.

@@

To compare the clusterings, we visualized cluster alignment with stacked barplots, and computed the adjusted rand index\cite{hubert1985} (ARI) and
Rosenberg's $v$-measure\cite{rosenberg07vmeasure} to evaluate similarity between the two
clusterings. Both measures range from 0 (no similarity) to 1 (identical), and respectively take a
contingency table and information-theoretic approach to measuring clustering similarity.

@@

$k$-means clustering performed on the 30 individual nonmotor symptoms found similar patterns to the
first clustering {(Table~\ref{tab:nms30}, heatmap in Figure~\ref{fig:nms30-hm}). Means of all
symptoms were found to differ across clusters except for PD onset and tremor.

@@

 The two clusterings were somewhat similar ($\text{ARI} = 0.46$; $\text{$v$-measure} = 0.43$). To
 help compare the alignment of the two clusterings, we denoted the four clusters in the domains
 clustering as $D_1, D_2, D_3, D_4$, and the clusters in the symptoms clustering as $S_1, S_2, S_3,
 S_4$.
Alignment of the clusters is visualized in Figure~\ref{fig:align}. While clusters $D_1/S_1$ and
$D_4/S_4$
aligned well with each other, main differences in the clusterings occurred in the mixing of
clusters 2 and 3; for example, $D_3$ (motor-dominant) was somewhat
evenly split among $S_1$ (mild) and $S_3$ (average).}

@@

Most variables had little to no correlation with PD duration (Supplementary Figure~\ref{fig:corr}). Variables
with the highest correlation include symptoms urinary frequency, swallowing, and drooling; domains
miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots for CISI Total,
Tremor, Anxiety, and Depression appear in Supplementary Figure~\ref{fig:long4}.

@@

Our longitudinal analysis gives more insight into the clusters found in the previous analyses.
According to Supplementary Figure~\ref{fig:corr}, most symptoms are uncorrelated with PD duration, especially
Mood/apathy symptoms (anxiety, depression) and tremor.
The differences in disease progression for each cluster can be seen by the corresponding graphs in
Figure~\ref{fig:long4}. Cluster 2 (Nonmotor-Dominant) starts at higher
scores for anxiety and depression, and actually decreases with increasing PD duration, thus indicating that these patients' subtype can be determined
early after PD onset from the depressive symptom score. Similarly, when examining cluster 3
(Motor-dominant), the mean tremor score is substantially higher from PD onset.
Interestingly, cluster 4 (Severe) generally starts at lower tremor and motor scores during disease,
but then rises sharply, exceeding other clusters. More evidence that tremor is a unique motor
symptom is located in Figure~\ref{fig:hc}, where it is the most distant symptom from all other
symptoms.

@@

\begin{suppfigure*}[p]
  \centering
  \includegraphics[width=0.8\linewidth]{long4-d.pdf}
  % TODO: Get rid of capital T in  CISI Total
  \caption{Selected symptoms plotted against PD duration. Smoothed loess curves for each cluster
  are drawn in their respective colors. The black curve is the curve for the entire population, and
the global mean score is marked with a dotted line.}
  \label{fig:long4}
\end{suppfigure*}
